Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2021-01-18 journal paper/content/sec-intro.tex
Minor edits for space.

The main high-level question we ask is:
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{Whether and how does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops
%--- when more data leads to more users, which leads to even more data, etc. ---
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

@@

\textbf{(i)} The most obvious decision rule maximizes the Bayesian-expected reward; we refer to it as \HardMax. We find that it is not conducive to adopting better algorithms: each principal's dominant strategy is to choose the greedy algorithm. Further,
%\HardMax is very sensitive to tie-breaking:
if the tie-breaking is probabilistically biased in favor of one principal, then this principal can always prevail in competition.
% has a simple ``winning strategy" no matter what the other principal does.

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax agents and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a sufficiently better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.

@@

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation: one that crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the traditional inverted-U relationship purely through the reputational consequences of exploration, rather than monetary costs/benefits thereof. This is in contrast to the inverted-U relationships established in prior work, which rely on the monetary aspects: how much can a firm invest into  R\&D and how much can it profit from the innovation.

@@

\xhdr{Numerical simulations.}
We consider a basic frequentist model (called the \emph{\ExptsModel}), where agents observe signals about the principals' past performance and
make their decisions naively,
%base their decisions on these signals alone,
without invoking any prior knowledge or beliefs. The performance signals are abstracted and aggregated as a scalar \emph{reputation score} for each principal, modeled as a sliding window average of its rewards. Thus, agents' decision rule depends only on the two reputation scores.
% We refer to this variant as the \emph{\ExptsModel}.%
%\footnote{In comparison, the theoretical results focus on another extreme, with Bayesian rationality and no performance signals.}
We refine and expand the theoretical results in several ways:

%\begin{itemize}
\textbf{(i)}
%We compare \HardMax and \HardMaxRandom decision rules.
We find that the greedy algorithm often wins under the \HardMax decision rule, with a strong evidence of the ``death spiral" effect mentioned earlier.
 As predicted by the theory, better algorithms prevail under \HardMaxRandom if the expected number of ``random" users is sufficiently large.
 %\asdelete{However, this effect is negligible for smaller parameter values.}

@@

\textbf{(iii)}
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance in a standalone bandit problem, \ie without competition, and we find that the most natural performance measure does not explain this phenomenon. We suggest a new, more nuanced performance measure that does.

>>>>

%The main high-level question we ask is:
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\asedit{We ask:}
\textbf{Whether and how does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops
%--- when more data leads to more users, which leads to even more data, etc. ---
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

@@

\textbf{(i)} The most obvious decision rule maximizes the Bayesian-expected reward; we call it \HardMax. We find that it is not conducive to adopting better algorithms: each principal's dominant strategy is to choose the greedy algorithm. Further,
%\HardMax is very sensitive to tie-breaking:
if the tie-breaking is probabilistically biased in favor of one principal, the latter 
%then this principal 
can always prevail in competition.
% has a simple ``winning strategy" no matter what the other principal does.

@@

\textbf{(iii)} We further soften the decision rule so that the selection probabilities vary smoothly in terms of principals' Bayesian-expected rewards;
%as a function of the difference between  principals' Bayesian-expected rewards;
We call it \SoftMaxRandom, a more realistic middle ground between \HardMax and random agents.%
\footnote{Alternatively, one can obtain a \SoftMaxRandom decision rule using a mixture of more ``basic" agent types that follow \HardMax unless the principal's Bayesian-expected rewards are too close to each other.}
In the most technical result of the paper, we find that a sufficiently better algorithm prevails under much weaker assumptions.
% on what constitutes a better algorithm.
%This is the most technical result of the paper.

@@

While traditional models of innovation study lab-based R\&D, we consider data-driven innovation: one that crucially depends on data generated by the firm's customers.
We focus on innovation in \emph{exploration technology} which systematically improves the firm's products, whereas prior work would define innovation as improvement in the products themselves. We recover the traditional inverted-U relationship purely through the reputational consequences of exploration.
\asdelete{rather than monetary costs/benefits thereof.}
This is in contrast to the inverted-U relationships established in prior work, which rely on the monetary aspects: how much can a firm invest into  R\&D and how much can it profit from the innovation.

@@

\xhdr{Numerical simulations.}
We consider a basic frequentist model (called the \emph{\ExptsModel}), where agents observe signals about the principals' past performance and
make their decisions naively,
%base their decisions on these signals alone,
without invoking any prior knowledge or beliefs. The performance signals are aggregated as a scalar \emph{reputation score} for each principal, modeled as a sliding window average of its rewards. Thus, agents' decision rule depends only on the two reputation scores.
% We refer to this variant as the \emph{\ExptsModel}.%
%\footnote{In comparison, the theoretical results focus on another extreme, with Bayesian rationality and no performance signals.}
We refine and expand the theoretical results in several ways:

%\begin{itemize}
\textbf{(i)}
%We compare \HardMax and \HardMaxRandom decision rules.
We find that the greedy algorithm often wins under the \HardMax decision rule, with a strong evidence of the ``death spiral" effect mentioned earlier. As predicted by the theory, better algorithms prevail under \HardMaxRandom
\asedit{with enough ``random" users.}
%if the expected number of ``random" users is sufficiently large.
 %\asdelete{However, this effect is negligible for smaller parameter values.}

@@

\textbf{(iii)}
However, the greedy algorithm is sometimes \emph{not} the best strategy under high levels of competition.\footnote{In our theoretical results on \HardMax, the greedy algorithm is always the best strategy, mainly because it is aware of the Bayesian prior (whereas in the simulations  the prior is not available).}
We revisit algorithms' performance in a standalone bandit problem, \ie without competition. We find that the most natural performance measure does not explain this phenomenon, and suggest a new, more nuanced \asedit{one} that does.
%performance measure that does.
